---
title: "One-way and paired t-tests examples, Cohen's d effect size, abd a table "
author: "Nicole Schmidt"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Take home lab for week 7


### Attach packages
```{r}

library(tidyverse)
library(here)
library(janitor)
library(effsize)
library(kableExtra)
```

### Background on flint water data
To practice paired and one sided t-tests, we'll use data collected from 271 resdidences in Flint, MI to explore the effect of a collection methods on observed lead concentrations (ppb)

**Background:** For a year and a half (summer 2014 - December 2015), city officials in Flint, MI directed residents to flush their faucets for 3-4 minutes before collecting samples for lead and copper corrosion testing. The guidlines were highly criticized by the EPA, as flushing could reduce Pb measured concentrations and imply lower exposure than resdients were actually experiencing. SO FUCKED

For more Flint residential testing and blood lead levels data: http://www.michigan.gov

### Read in Flint water data
```{r}

flint_pb <-read_csv("flint_water.csv") %>% 
  clean_names() %>% 
  rename(pb_immediate = pb_bottle_1_ppb_first_draw,
         pb_2min = pb_bottle_2_ppb_2_mins_flushing)

```

We are asked: is there a significant difference between lead concentrations immediately sampled and after 2 minutes of flushing?

First, we'll look at the distributions and sample sizes:
- With 271 observations each (note: there are 5 NA values in each sample), we have enough to say that by Central Limit Theorem we know the sampling distribution of means will be normal regardless of underlying population, so comparing means using a t-test is OK if I think means is a useful metric of comparison. 
- We should still always LOOK at the data, anyway:

```{r}

#Histograms
ggplot(flint_pb, aes(x = pb_immediate))+
  geom_histogram()

```


```{r}
ggplot(flint_pb, aes(x = pb_2min))+
geom_histogram()
```

```{r}
#QQ plots
ggplot(flint_pb, aes(sample = pb_immediate))+
  geom_qq()

```

```{r}
ggplot(flint_pb, aes(sample = pb_2min))+
  geom_qq()
```

There are NOT normally distributed; we will say we want to compare means, and can do that comfortabley because of the CLT (sampling distribution of the means will be normal)

We should also look at some statistics for comparison:

```{r}
# Sample size (non-NA)
n_immediate <- sum(!is.na(flint_pb$pb_immediate))
n_flushed <- sum(is.na(flint_pb$pb_2min))

# Means (immediate v. 2 min flushing):
mean_immediate <- mean(flint_pb$pb_immediate, na.rm = TRUE)
mean_flushed <- mean(flint_pb$pb_2min, na.rm = TRUE)

# Medians (Immediate v. 2 min flushing):
median_immediate <- median(flint_pb$pb_immediate, na.rm = TRUE)
median_flushed <- median(flint_pb$pb_2min, na.rm = TRUE)

#Standard deviations (immediate v. 2 min flushing):
sd.immediate <- sd(flint_pb$pb_immediate, na.rm = TRUE)
sd.flushed <- sd(flint_pb$pb_2min, na.rm = TRUE)

# Lead concentrations measured in immediately collected samples are much higher than those in water collected after 2 min flushing. 
```
Return the values for the sample sizes and summary statistics in the code chunk above to see them in the Console. How do the sample means of lead concentration compare for immediate vs. 2 min flushed water samples?

### Are observations paired or unpaired? 
These observations are reported for each *household.* Does it make sense to compare the immediate and post-flushing observations across different households?

**No.** It makes sense to recognize that when we compare values, we should be comparing the immediate and post-flushing lead concentration differences at each house (e.g., each observation in the 'immediate' sample is associated with one and only one observation in the '2 min flushing' sample).

When that is the case, data are called **paired**, and we will perform a **paired t-test** to answer: "Is there a significant difference in lead concentration in immediately collected tap, and in water after 2-min of flushing?"

**Null hypothesis:** The mean lead concentration is the same in water collected immediately, and water collected after 2-min flushing. 

**Alternative hypotheses:** The mean lead concentration is NOT the same in water collected immediately, and water after 2-min flushing. 

### Two-sample, two-sided paired t-test
to answer that question, we'll perform a two-sided, two-sample paired t-test. Breaking it down:

- **Two-sided** becasue we're not asking "Is A greater than B" or is "B less than A", we're just asking if they're different in either direction

- **Two-sample** because we're comparing means of two samples

- **Paired** because each observation in sample A is associated with one and only one observation in sample B

Perform the test by inputting the sample vectors, and adding argument paired = TRUE:

```{r}
my_flint_test <- t.test(flint_pb$pb_immediate, flint_pb$pb_2min, paired = TRUE)
```

**Example statement of test outcome:**
"Mean lead concentration (ppb) measured in immediately collected water samples (10.65 ± 21.56, n = 271) differed significantly from lead in water collected after 2 minutes of flushing (3.66 ± 10.54, n = 271) by a paired two-sample t-test (t(270)=6.37, p<0.001)."

**Note:** Usually when a p-value is really tiny, p < 0.001 is sufficient. But ask: Why is the way I added that in the statement above *not* best practice for reproducibility? 

####Two Sample, ONE-sided paired t-test:
What if our question isn't "do the means differ," but instead "are mean lead concentrations in water after 2-min flushing *LESS* than in immediately sampled water?"

Then we are implying directionality, and would want to perform a one sided test. We add directionality to t.test() by including the argument alternatice = "greater" or alternative = "less", depending on the order that we add our samples in the function.

If i want to test: Is mean of A greater than mean of B? Then my code would be:
t.test(A,B, alternative = "greater')

If I want to test: Is the mean of B *less* than mean of A? (note that that is the same as the question above, just asked differently) than my code would be:
t.test(B,A, alternative = "less")

So be careful of order of inputs when you're doing a one-tailed t-test!

Our question is: "Are mean lead concentration in water after 2-min flushing *LESS* than in immediately sampled water?"

**Null hypotheses:** The mean lead concentration in flushed samples *is not lower* than the mean for immediately sampled water.

**Alternative hypotheses:** The mean lead concentration in flushed samples *is lower* than the mean for immediately sampled water. 

Perform a one-sided, two-sample paired t-test:
```{r}

flushed_less_ttest <- t.test(flint_pb$pb_immediate,
                             flint_pb$pb_2min,
                             paired = TRUE,
                             alternative = "less")
flushed_less_ttest


```

???IN LINE REFERENCING???

### Cohen's d effect size
remember the p value is not enough. Here, we'll use Cohen;s d effect size to report a more meaningful metric of differences between group means.

We could write our own function to calculate the effect size

```{r}

#Creating a function called 'calc_d' to calculate Cohen's d effect size 
# Here, a and b will ne our sample vectors

calc_d <- function(a,b) {
  sd_pooled <- sqrt((sd(a,na.rm = TRUE)^2 + sd(b,na.rm = TRUE)^2)/2) #sd pooled
                    x <- (mean(a,na.rm = TRUE) - mean(b, na.rm = TRUE))/sd_pooled #complete equation
                    return(x)
}

# Then apply that function to our samples

flint_d_myfunction <- calc_d(flint_pb$pb_immediate, flint_pb$pb_2min)
# d = 0.41 (moderate effect size)
```

Or we could use the exisdting effsize::cohen.d() function instead of creating our own. Let's use it here to check that results match for our function and the cohen.d function

```{r}
flint_d <- effsize::cohen.d(flint_pb$pb_immediate, flint_pb$pb_2min, na.rm = TRUE)
# Same returned! Cohen's d = 0.41 (moderate effect size)
```

Then in our report we would want to include the actual means of the samples, and the effect size, possibly the confidence interval for each, and then the *least interesting thing* should be statement of significant associate with our t-test.

### Making a table with kableExtra
**Note:** There are a million ways to make tables. Some useful packages are:
- DT
- kable
- kableExtra
- flextable
- gt

Heres an example using kableextra
using a subset from built in R datas 'trees'
original data can be seen with View(trees)
Make a subset then create a table with kableExtra

```{r}

#Make the subset (keep only rows 1-5):
trees_sub <- trees %>% 
  dplyr::slice(1:5)

# Make the table of the subset
trees_sub %>% 
  kable(col.names = c("Diameter (inches)",
                      "Height (ft)",
                      "Volume (cubic feet")) %>% 
  kable_styling(bootstrap_options = "striped",
                full_width = F,
                position = "left") %>% 
  add_header_above(c("Black cherry tree metrics" = 3))
```

Lots of other customization options!
